---
title: "Divvy Bike Share User Analysis"
author: "Ekeoma E. AGU"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup environment
The packages for data cleaning and analysis are loaded

```{r}
library(tidyverse)
library(lubridate)
library(skimr)
library(naniar)

```

## Get data
Uploading all Divvy trip data from October 2021 - September 2022 (the past 12 months)

```{r}
# Get files from working directory
data_files <- list.files(file.path(getwd(), "data"))

for(i in 1:length(data_files)){
  assign(paste0("month_", i),
         read.csv(paste0(getwd(), "/data/", data_files[i])))
}
```

## Review and Join dataframes to into one
Here we verify that the column naming convention and data types are consistent across all twelve data files to ensure the accuracy of the data set. 

```{r}
glimpse(month_1)
glimpse(month_2)
glimpse(month_3)
glimpse(month_4)
glimpse(month_5)
glimpse(month_6)
glimpse(month_7)
glimpse(month_8)
glimpse(month_9)
glimpse(month_10)
glimpse(month_11)
glimpse(month_12)
```

#### Combine all dataframes to one
Now that we have verified that the field names are consistent across all twelve csv files, let's combine the data to one dataframe and view the first few rows.
```{r}
combined_trips <- bind_rows(month_1, month_2, month_3, month_4,
                        month_5, month_6, month_7, month_8,
                        month_9, month_10, month_11, month_12)

head(combined_trips)
```
### Check for duplicate records
Next we will check to verify that there are no duplicate records
```{r}
sum(duplicated(combined_trips))
```
This output shows that there are no duplicate values, each row of the trip record represent a unique trip.