---
title: "Divvy Bike Share User Analysis"
author: "Ekeoma E. AGU"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup environment
The packages for data cleaning and analysis are loaded

```{r}
library(tidyverse)
library(lubridate)
library(skimr)
library(naniar)

```

## Get data
Uploading all Divvy trip data from October 2021 - September 2022 (the past 12 months)

```{r}
# Get files from working directory
data_files <- list.files(file.path(getwd(), "data"))

for(i in 1:length(data_files)){
  assign(paste0("month_", i),
         read.csv(paste0(getwd(), "/data/", data_files[i])))
}
```

## Review and Join dataframes to into one
Here we verify that the column naming convention and data types are consistent across all twelve data files to ensure the accuracy of the data set. 

```{r}
glimpse(month_1)
glimpse(month_2)
glimpse(month_3)
glimpse(month_4)
glimpse(month_5)
glimpse(month_6)
glimpse(month_7)
glimpse(month_8)
glimpse(month_9)
glimpse(month_10)
glimpse(month_11)
glimpse(month_12)
```

#### Combine all dataframes to one
Now that we have verified that the field names are consistent across all twelve csv files, let's combine the data to one dataframe and view the first few rows.
```{r}
combined_trips <- bind_rows(month_1, month_2, month_3, month_4,
                        month_5, month_6, month_7, month_8,
                        month_9, month_10, month_11, month_12)

head(combined_trips)
```
### Check for duplicate records
Next we will check to verify that there are no duplicate records
```{r}
sum(duplicated(combined_trips))
```
This output shows that there are no duplicate values, each row of the trip record represent a unique trip.

Check shape of data frame to understand the total number of trip records collected over the period of interest.
```{r}
print(dim(combined_trips))

number_of_records = dim(combined_trips)[1]
```

## Clean Data and Prepare for analysis
#### Rename ride_id and rideable_type columns with a more descriptive names

```{r}
combined_trips <- combined_trips %>% 
  rename(trip_id=ride_id, bike_id=rideable_type)

glimpse(combined_trips)
```

#### Get a general overview of the data frame
Get summary statistics and check for missing values as well as detect attributes with empty values.
```{r}
skim_without_charts(combined_trips)
```

#### Remove records without start or end station details

```{r}
all_valid_trips <- combined_trips %>% 
  filter(end_station_id !="" & start_station_id !="")

# Review the summary detail
skim_without_charts(all_valid_trips)
```


#### What is the total number of valid trips
```{r}
# Number of valid trips
num_of_valid_trips = nrow(all_valid_trips)

num_of_valid_trips
```

#### What is the number of trips with missing details
```{r}
num_of_invalid_trips = number_of_records - num_of_valid_trips

num_of_invalid_trips
```

#### What percentage of the trip records were invalid
```{r}
percent_of_invalid_trips = (num_of_invalid_trips/number_of_records)*100

percent_of_invalid_trips
```

### Create the trip_duration and week_day fields

```{r}

all_valid_trips <- all_valid_trips %>% 
  mutate(week_day = wday(started_at, label = TRUE), wday = wday(started_at))

#all_valid_trips$week_day <- format(as.Date(all_valid_trips$started_at), "%A")
all_valid_trips$trip_duration <-  round(as.numeric(difftime(all_valid_trips$ended_at, all_valid_trips$started_at))/60,2) 

head(all_valid_trips)
```

#### What is the average trip duration?

```{r}
with(all_valid_trips, round(mean(trip_duration),1))
```

#### What is the proportion of user types?

```{r}
all_valid_trips |> 
  group_by(member_casual) |> 
  summarise(number_of_trips = n()) |> 
  ggplot(aes(x=member_casual, y=number_of_trips, fill = member_casual)) +
  geom_bar(stat = "identity") +
  theme_minimal()+
  ggtitle("Proportion of User Types by Number of Trips")+
  xlab("")+ylab("")+theme(axis.line = element_line(color='black'))+
  geom_text(aes(label=paste0(round((number_of_trips/1000000),2),"M ",
                             paste0("(",round((number_of_trips/num_of_valid_trips)*100,1),"%",")"))))
```

#### What is the most used bike types by membership types?

```{r}
all_valid_trips %>% 
  group_by(member_casual, bike_id) %>% 
  summarise(avg_trip_duration = mean(trip_duration), number_of_trips = n(), .groups="keep") %>%
  ggplot(aes(x=reorder(bike_id, -number_of_trips), y=number_of_trips, fill=member_casual))+
  geom_bar(stat = "identity") +
  theme_minimal()+
  ggtitle("Most used Bikes by User Type")+
  xlab("")+ylab("")+theme(axis.line = element_line(color='black'))+
  geom_text(aes(label=paste0(round((number_of_trips/1000),0),"K")))+
  coord_flip()+
  facet_wrap(~member_casual)
```

#### What is the average trip duration by membership type?

```{r}
all_valid_trips %>% 
  group_by(member_casual, bike_id) %>% 
  summarise(avg_trip_duration = mean(trip_duration), number_of_trips = n(), .groups = "keep") %>% 
  ggplot(aes(x=member_casual, y=avg_trip_duration, fill=bike_id))+
  geom_bar(stat = "identity")+
  theme_minimal()+
  ggtitle("Trip Duration by User type per Bike type")+
  xlab("")+ylab("")+theme(axis.line = element_line(color='black'))+
  geom_text(aes(label=round(avg_trip_duration,1)))+
  facet_wrap(~bike_id)
```

#### Which week day do casual riders have peak bike usage?

```{r}
all_valid_trips %>% 
  filter(member_casual=="casual") |> 
  group_by(week_day, wday) %>% 
  summarise(avg_trip_duration = mean(trip_duration), number_of_trips = n()) %>% 
  ggplot(aes(x=week_day, y=number_of_trips)) +
  geom_bar(stat = "identity", fill = "sky blue") +
  theme_minimal()+
  ggtitle("Total trips per Weekday by Casual riders")+
  xlab("")+ylab("")+theme(axis.line = element_line(color='black'))
```

#### Which week days do annual members have peak bike usage?

```{r}
all_valid_trips %>%
  filter(member_casual=="member") %>%
  group_by(week_day, wday) %>%
  summarise(avg_trip_duration = mean(trip_duration), number_of_trips = n()) %>%
  ggplot(aes(x=week_day, y=number_of_trips)) +
  geom_bar(stat = "identity", fill = "#D5EA67") +
  theme_minimal()+
  ggtitle("Total trips per Weekday by Annual members")+
  xlab("")+ylab("")+theme(axis.line = element_line(color='black'))
```

#### Which week days do the membership types record highest average trip duration?

```{r}
all_valid_trips %>% 
  group_by(week_day, wday, member_casual) %>% 
  summarise(avg_trip_duration = mean(trip_duration), number_of_trips = n()) %>% 
  ggplot(aes(x=week_day, y=avg_trip_duration, fill=member_casual)) +
  geom_bar(stat = "identity") +
  theme_minimal()+
  ggtitle("Avg. Trip duration per Weekday")+
  xlab("")+ylab("")+theme(axis.line = element_line(color='black'))
```

#### What are the most used routes by the membership types?
To evaluate this measure, there is need to create the route column

```{r}
all_valid_trips$route <- paste(all_valid_trips$start_station_name, "_to_", all_valid_trips$end_station_name)

head(all_valid_trips$route)
```

#### Most used routes by casual riders

```{r}
all_valid_trips %>% 
  filter(member_casual == "casual") |> 
  group_by(route) %>% 
  summarise(avg_trip_duration = mean(trip_duration), number_of_trips = n(), .groups = "keep") %>% 
  arrange(-number_of_trips) |> 
    head(5) |> 
  ggplot(aes(x=reorder(route, +number_of_trips), y=number_of_trips))+
  geom_bar(stat = "identity", fill="sky blue")+
  theme_minimal()+
  geom_text(aes(label=paste0(round((number_of_trips/1000),1),"K")))+
  xlab("")+ylab("")+theme(axis.line = element_line(color='black'))+
  coord_flip()
```

#### Most used routes by annual members

```{r}
all_valid_trips %>% 
  filter(member_casual == "member") %>% 
  group_by(route) %>% 
  summarise(avg_trip_duration = mean(trip_duration), number_of_trips = n(), .groups = "keep") %>% 
  arrange(-number_of_trips) %>% 
    head(5) %>% 
  ggplot(aes(x=reorder(route, +number_of_trips), y=number_of_trips))+
  geom_bar(stat = "identity", fill="sky blue")+
  theme_minimal()+
  geom_text(aes(label=paste0(round((number_of_trips/1000),1),"K")))+
  xlab("")+ylab("")+theme(axis.line = element_line(color='black'))+
  coord_flip()
```